Palmer Archipelago (Antarctica) penguins: an alternative to Edgar Anderson's iris data

### 2020-06-25 manuscript/manuscript_text.Rmd
add where else to get penguins

In 1935, American botanist Edgar Anderson measured petal and sepal dimensions (length and width) for 50 samples from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson]. The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data (....SOMETHING HERE) make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, hypothesis testing, regression, machine learning, and more. Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R*, Python's *scikit-learn* machine learning library, and the SAS *Sashelp* library, further facilitating its widespread use. As a result, eighty-five years after the data were initially published, Anderson’s Iris measurements are ubiquitous in statistics and data science materials.

@@

In 1936, one year after publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A. Fisher in *The Annals of Eugenics* [@fisher]. Regardless of Anderson's initial motivation, the dataset today inextricably linked to Fisher’s eugenics research and even commonly (if unfairly) referred to as "*Fisher's* iris data" (ref). Due to its history in - and association with - Fisher's eugenics research, there is momentum to replace Iris with alternative datasets that offer similar teaching and learning opportunities.

Considering *Iris* usage, we established the following criteria for a suitable alternative dataset:

- Available by an appropriate license, ideally CC0 (no rights reserved)
- Feature intuitive subjects and variables understandable to learners across fields
- Manageable, but not trivial, size
- Real-world observations for improved understanding and engagement [@cafe]
- Provide similar opportunities for teaching data science skills and statistical methods

Here, we offer an alternative to Anderson's irises that largely satisfies those criteria: an approachable and charismatic dataset containing real-world size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica. Comparable size, characteristics and learning opportunities make the penguins data uniquely transferable for educators currently using *Iris*. By highlighting similarities in data structure and analyses, as well as several examples for which the penguins are *not* a good alternative, we demonstrate that the penguins can replace irises for many use cases.

@@

As a direct alternative to Anderson's *Iris* data, we recommend using the curated `penguins` object in the *palmerpenguins* data package. The `penguins` data is a subset of `penguins_raw` retaining all 344 rows (individual penguins) but reduced to the following seven variables:

@@

The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`, percent missing is `r round(100*sum(is.na(penguins))/(7*344),2)` %) indicated by `NA` throughout. In addition to reduced variables, the curated `penguins` subset is minimally updated from `penguins_raw` as follows:

@@

We feel that the `penguins` data is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`. Comparisons presented here are representative examples, and are not exhaustive.

@@

### Correlation

@@

@@

### Classification?

### Other comparative analyses?

@@

- t-tests?
- ANOVAs?
- MANOVAs?

## Suggested activities

### Activity 1: Predict penguin body mass

@@

## Random other stuff

@@

>>>>

In 1935, American botanist Edgar Anderson measured petal and sepal dimensions (length and width) for 50 samples from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson]. The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, classification and more. Anderson's *Iris* dataset is even built into a number of software packages, including in the auto-installed *datasets* package in *R* (R Core Team), Python's *scikit-learn* machine learning library [@scikit-learn], and the SAS *Sashelp* library, further facilitating its widespread use. As a result, eighty-five years after the data were initially published, Anderson’s Iris measurements are ubiquitous in statistics and data science education.

@@

In 1936, one year after publication in the *Bulletin of the American Iris Society*, Anderson’s *Iris* measurements were further analyzed and published by eugenicist and statistician Ronald A. Fisher in *The Annals of Eugenics* [@fisher]. Regardless of Anderson's initial motivation, the dataset today remains inextricably linked to Fisher’s eugenics research and even commonly (if unfairly) referred to as "*Fisher's* iris data" (ref). Due to its history in - and strong association with - Fisher's eugenics research, there is momentum to replace the *Iris* data with alternative datasets that offer similar teaching and learning opportunities.

Considering *Iris* data usage, we established the following criteria for a suitable alternative dataset:

- Available by an appropriate license, ideally CC0 (no rights reserved)
- Feature intuitive subjects and variables understandable to learners across fields
- Manageable - but not trivial - in size
- Real-world (not manufactured) data for improved understanding and engagement [@cafe]
- Provide similar opportunities for teaching data science skills and statistical methods

Here, we offer an alternative to Anderson's *Iris* data that largely satisfies those criteria: an approachable and charismatic dataset containing real-world size measurements for three penguin species observed on islands in Palmer Archipelago, Antarctica. Comparable size, characteristics and learning opportunities make the penguins data uniquely transferable for educators currently using *Iris*. By highlighting similarities in data structure and analyses, as well as several examples for which the penguins are *not* a good alternative, we demonstrate that the penguins can replace irises for many use cases.

@@

As a direct alternative to Anderson's *Iris* data, we recommend using the curated `penguins` object in the **palmerpenguins** *R* data package. The `penguins` data is a subset of `penguins_raw` retaining all 344 rows (individual penguins) but reduced to the following seven variables:

@@

The `penguins` subset contains `r sum(complete.cases(penguins))` complete cases, with missing values (n~missing~ = `r sum(is.na(penguins))`, or `r round(100*sum(is.na(penguins))/(7*344),2)` % of all values) indicated by `NA` throughout. In addition to reduced variables, the curated `penguins` subset is minimally updated from `penguins_raw` as follows:

@@

We feel that the `penguins` data is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`. Comparisons presented here are representative examples, and are not exhaustive. Additional examples and tutorials are shared in the [user-contributed examples](https://allisonhorst.github.io/palmerpenguins/articles/articles/user_contributions.html) article on the package website.

@@

### Linear correlation & regression opportunities

@@

More obvious clustering (by penguin species) exists when comparing bill length and depth (Figure X), and the relationship also provides an opportunity to introduce & explore Simpson's Paradox.

@@

@@

### Principle components analysis

### Clustering (e.g. k-means?)

### Classification (Alison Hill blog post?)

## Penguin data access for non-R users

The Palmer Archipelago penguins data are available to non-R users by direct import from the Environmental Data Initiative, from GitHub, and as a stored dataset in existing packages. Below we share several current ways to access the data (as of July 2020 - we expect availability to expand in the future).

### Download or import directly from Environmental Data Initiative

The [Environmental Data Initiative](https://environmentaldatainitiative.org/) is "an NSF-funded project, actively promoting and enabling curation and re-use of environmental data." The raw data for each penguin species (Adélie, Gentoo and Chinstrap) are stored in a separate data file at the following links (see full citations in Citations):

- Adélie data: [KNB-LTER Data Package 219.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5)
- Gentoo data: [KNB-LTER Data Package 220.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5)
- Chinstrap data: [KNB-LTER Data Package 221.5](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6)

The links above contain all metadata information, and provides options to directly download the data as a comma-separated value (CSV) file *or* import for use directly using code provided (e.g. for R, Python, SAS, MatLab, or SPSS).

### Download CSV from the palmerpenguins GitHub repo

A CSV containing curated data for all three species (accessed directly from EDI) exists in the 'data-raw' folder in the palmerpenguins repo on GitHub. GitHub users can fork / clone the repo, and will then have the cleaned CSV containing all 3 penguin species.

### The penguins in other packages & languages

The Palmer Station penguin data has been added to existing libraries, including (non-exhaustive):

#### Python modules

- Seaborn
- blah blah

See examples for working with penguins in Python here:

#### Kaggle

- here

#### Elsewhere?

## Conclusion

@@

# Random other stuff

@@

A Web of Science search reveals that Fisher's 1936 paper has been cited in 7,258 publications (as of June 2020), most often in journals publishing in fields of Computer Science (57% of 7,258 citing references), Mathematics (49%), and Engineering (29%) - most of which we expect are crediting Anderson's *Iris* data, and not Fisher's eugenics work.

# End random other stuff
